A generalized definition of multidimensional item response theory parameters

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd

```{r setup, include=FALSE}
# Includes:

library(knitr)
library(patchwork)

ROOT_DIR <- "../.."
DOC_DIR <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png"),
  fig.pos = "!H"
)

@@

```{r write-bib, cache=TRUE}
# automatically create a bib database for R packages
write_bib(
  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
  'www/packages.bib'
)
```

```{r sources, cache=FALSE, include=FALSE}
source("R/Formulae.R", encoding = 'UTF-8')
```

@@

@reckase_difficulty_1985 argues for a need to take dimensionality into account
when characterizing the statistical properties of multidimensional items.
As such, he outlines a general procedure for determining their difficulty.
Following this procedure, he also derives
a *multidimensional difficulty* parameter for M2PL items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating locus of the item.
The second one, the direction of the vector from the origin to that same locus.
Following a similar rationale, @reckase_discriminating_1991 derive a
*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
for that model.

@@

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994,
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the M2PL model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next we discuss the properties of the indices,
propose a graphical representation method,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

```{r libraries, message=FALSE}
```

```{r sources, message=FALSE}
```

@@

## Example with empirical data

```{r reset-root-wd}
# Necessary for knitr to find the source file:
opts_knit$set(root.dir = ROOT_DIR)
```

@@

```{r libraries, message=FALSE}
```

```{r sources, message=FALSE}
```

```{r read-items, message=FALSE}
```

```{r orthogonal-params}
```

```{r oblique-params}
```

```{r params-table}
```

```{r compose-empirical-items-table}
```

```{r set-flextable-wd, cache=FALSE}
```

@@

```{r empirical-items-table-out, cache=FALSE, tab.id="empirical-items-table-out", tab.cap="Item parameters for the empirical example"}
item_params_output
```

>>>>

```{r setup}
#| include: false

# Includes:

library(knitr)
library(patchwork)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png")
)

@@

```{r write-bib}
#| cache: true

# automatically create a bib database for R packages
write_bib(
  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
  'www/packages.bib'
)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Formulae.R",  encoding = 'UTF-8')
source("R/Constants.R", encoding = 'UTF-8')
```

@@

@reckase_difficulty_1985 argues for a need to take dimensionality into account
when characterizing the statistical properties of multidimensional items.
As such, he outlines a general procedure for determining their difficulty.
Following this procedure, he also derives
a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating locus of the item.
The second one, the direction of the vector from the origin to that same locus.
Following a similar rationale, @reckase_discriminating_1991 derive a
*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
for that model.

@@

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994,
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next we discuss the properties of the indices,
propose a graphical representation method,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

```{r libraries}
```

```{r sources}
```

@@

## Examples with empirical data

```{r reset-root-wd}
#| cache: false

# Necessary for knitr to find the source file:
opts_knit$set(root.dir = ROOT_DIR)
```

@@

```{r libraries}
```

```{r sources}
```

```{r general-constants}
```

```{r set-cov-matrix-Reckase2009}
# Version in Reckase, 2009, p. 153; this represents a usual 3D case
COV_MATRIX <- matrix(
  c(
    1.210, .297, 1.232,
     .297, .810,  .252,
    1.232, .252, 1.960
  ),
  nrow = 3
)
# (see also Reckase, 2009, p. 183 for a "close-to-1D" case).
```

```{r read-items}
```

```{r compute-orthogonal-params}
```

```{r compute-oblique-params}
```

```{r create-params-table}
```

```{r compose-output-table}
```

```{r set-flextable-wd}
#| cache: false
```

```{r empirical-table-Reckase2009-out}
#| cache:   false
#| tab.id:  empirical-table-Reckase2009-out
#| tab.cap: Item parameters for the empirical example
item_params_output
```

```{r set-cov-matrix-Tezza2018}
# See Tezza et al., 2018, p. 927:
# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
# > foi de aproximadamente 0,4.
COV_MATRIX <- diag(4)
COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
```

```{r reset-root-wd}
#| cache: false
```

```{r read-items-Tezza2018}
items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
```

```{r compute-orthogonal-params}
```

```{r compare-orthogonal-params}
```

<!-- Computed item parameters in the orthonormal space match the ones in given
in Tezza 2018, table 5 -->

```{r compute-oblique-params}
```

```{r create-params-table}
```

```{r compose-output-table}
```

```{r set-flextable-wd}
#| cache: false
```

@@

```{r empirical-table-Tezza2018-out}
#| cache:   false
#| tab.id:  empirical-table-Tezza2018-out
#| tab.cap: Item parameters for the empirical example
item_params_output
```
